Suitability of the local-maximum entropy approximation under the Material Point Method framework for dynamic problems. \thanks{Funding: This study was funded by Agustín de Betancourt Foundation (grant number 262390106114).}

### 2020-02-25 main.tex
Cambios 25-2-2020

\section{Introduction}
\label{intro}
Since the proposal of the Material Point Method (\textbf{MPM}) by Sulsky {\it
  et al.} (1994)\cite{Sulsky1994}, as a generalization of
a particle-in-cell method \cite{Harlow1956} and a fluid implicit
particle (FLIP) method \cite{Brackbill1986}, MPM has been used to
simulate dynamic problems. However, in the simulations made with the
originalMPM, there are numerical noises \cite{Tran2019e} when it is faced to solve
classic dynamic benchmarks like the one proposed by Dyka \& Ingel
(1995)\cite{Dyka1995}.

The aim of this paper is to mitigate this spurious oscillations by the
employ of the maximum-entropy (or local \textit{max-ent}) shape
function under the MPM framework. First introduced by
\cite{Arroyo2006}, it belongs to the class of convex 
approximation schemes and provides a seamless transition between
finite elements (\textbf{FE}) and mesh-free interpolations. The
approximation scheme is based on a compromise between minimizing the
width of the shape function support and maximizing the information
entropy of the approximation. The local \textit{max-ent} approximation
may be regarded as a regularization, or \textit{thermalization}, of
Delaunay triangulation which effectively resolves the degenerate cases
resulting from the lack of uniqueness of the triangulation. Local
\textit{max-ent} basis functions possess many desirable properties for
meshfree algorithms. First of all, they are entirely defined by the
nodal set and the domain of analysis. They are also non-negative, satisfy the partition of unity
property, and provide an exact approximation for affine functions \cite{Arroyo2006}.

This approximation scheme has been proof to have a good performance
under the dynamic regime by other researchers like Navas {\it et al.}
(2018)\cite{Navas2018a} and Li {\it et al.} (2012)\cite{Li2012} for
Optimal Transportation Meshfree (\textbf{OTM}) method.

@@

\section{Notation}
\label{sec:Notation}
In what follows, we will adopt the following convention:
\begin{table}[h]
  \centering
  \caption{Physical variables involved in the problem}
  \label{tab:notation-table}
  \begin{tabular}{c c c}
    $\rho$ & Density field & Scalar \\
    $\vect{a}$ & Acceleration field & First order tensor \\
    $\vect{v}$ & Velocity field & First order tensor \\
    $\vect{u}$ & Displacement field & Second order tensor \\
    $\tens{\sigma}$ & Cauchy stress tensor & Second order tensor \\
    $\tens{\varepsilon}$ & Cauchy strain tensor & Second order tensor \\
    $\tens{D}$ & Constitutive tensor & Fourth order tensor\\
  \end{tabular}
\end{table}

The subscript $p$ is used to define a particle variable. While the
subscript $I$ is reserved in this notation for denoting nodal
variables.

The superscript $\psi$ involves a virtual magnitude.

$\dot{\qed}$ 
where $\Div{\qed}$ denotes the divergence operator.

where $\GradS{\qed}$ denotes the symmetric part of the gradient
operation.

$\varGamma_d$  $\varGamma_n$
\section{Derivation of the MPM}
\label{sec:derivation-mpm}
In this section, we provide an overview of the standard explicit MPM
algorithm as presented by \cite{Sulsky1994}. The method has three main
steps: a variational recovery process to project particle data to the
nodes of the grid, a Eulerian step where balance of momentum is
compute by solving internal and external forces in the nodes, and
finally the advection of the Lagrangian particles. We will first give
a introduction to the governing equations and then follow by an
algorithmic description of the process.

\subsection{Governing equations }
\label{sec:governing-equations}
In the MPM approach a continuum is considered. Let define a
$\varOmega$ domain occupied by an elastic body, and $\partial
\varOmega$ the boundary of the domain, where $\partial
\varOmega = \Gamma_d \bigcup \Gamma_n$ and $\Gamma_d \cap  \Gamma_n =
\emptyset$. The behaviour of the continuum can be described by a set
of governing equations which are (i) the balance of momentum equation
\begin{equation}
  \label{eq:Balance-momentum}
\rho \vect{a} = \Div{\tens{\sigma}} + \rho \vect{b},
\end{equation}
next (ii) the compatibility equation with ensure that the velocity
field is compatible with the Cauchy strain field,
\begin{equation}
  \label{eq:Compatibility-equation}
  \dot{\tens{\varepsilon}} = \GradS{\vect{v}} = \frac{1}{2}\left(
    \Grad{\vect{v}} +  \GradT{\vect{v}} \right),
\end{equation}
later (iii) the constitutive equation, which relates the rate of Cauchy strain
tensor with the rate of Cauchy stress tensor,
\begin{equation}
  \label{eq:Constitutive-equation}
\dot{\tens{\sigma}} = \tens{D} \colon \dot{\tens{\varepsilon}},
\end{equation}
finally we introduce (iv) the mass conservation, which if obtained by
setting to zero the total derivative of the density field,
\begin{equation}
  \label{eq:Rho-material-derivative}
  \frac{D \rho}{D t} = \dot{\rho} + \rho \Grad{\vect{v}} = 0.
\end{equation}

@@

\subsection{Galerkin procedure}
\label{sec:Galerkin-procedure}
In order to arrive to a finite set of equations, in contrast with the
FEM, in the MPM a double discretization procedure is performed as we will
describe here below. First, the continuum domain
$\Omega$ is discretized with a finite sum of material points (in the
following particles), each one
represent a part of the discretized domain $\varOmega_p \subset
\varOmega$ with $p = 1,2\ldots ,N_p$ where $N_p$ is the number of
particles. The material point $\vec{x}_p$ is defined at the centroid
of each $\Omega_p$, figure \ref{fig:MPM-discretization}.
\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/Discretization_MPM}
  }
  \caption{MPM discretization.}
  \label{fig:MPM-discretization}
\end{figure}

Each material point is assigned with initial values of position,
velocity, mass, volume and stress denoted by $\vec{x}_p$,
$\vec{v}_p$, $m_p$,  $V_p$ and $\tens{\sigma}_p$, but also the
virtual field $\vect{v}^{\psi}_{p}$. So employing the definition of the
material integral, where we recover the Riemann integral definition as
an addition  of a finite set of points. Therefore individual terms in
\eqref{eq:BalanceMomentum_wf} are solved as follows. 
\begin{itemize}
\item Acceleration forces :
\begin{equation}
    \label{eq:particle_acceleration_forces}
    \Integral{3}{\rho\ \vec{a}\ \vect{v}^{\psi}} =& \sum^{N_p}_{p}
   \dot{\vec{v}}_{p}\ \vect{v}^{\psi}_{p}\ m_p.
  \end{equation}\\
\item Internal forces :
  \begin{equation}
    \label{eq:particle_internal_forces}
    \Integral{3}{\tens{\sigma}\ \colon \Grad{\vec{\psi}}} = \sum^{N_p}_{p}
   \tens{\sigma}_{p}\ \colon \Grad{\tens{\psi}_{p}}\ V_p.
  \end{equation}\\
\item Body forces :
\begin{equation}
  \label{eq:particle_body_forces}
  \Integral{3}{\rho\ \vec{b}\ \vect{v}^{\psi} } = \sum^{N_p}_{p}
  \vec{b}_{p}\ \vect{v}^{\psi}_p\ m_p.
\end{equation}\\
\item Loads :
\begin{equation}
  \begin{aligned}
    \label{eq:particle_load_forces}
    \Integral{2}{\vec{t}\ \vect{v}^{\psi}} = \Integral{2}{\rho\
      \vec{t}^s\ \vect{v}^{\psi}} = \sum^{N_p}_{p} \vec{t}^s_{p}\ \vect{v}^{\psi}_{p}\ h^{-1}\ m_p ,
  \end{aligned} 
\end{equation}
\end{itemize}
where $h$ is the thickness of the continuum in a 2D case. Here is
where the second discretization procedure appears. A background mesh
composed by a finite set of grid points with coordinates $\vec{x}_I\
, I = 1,2\ldots ,N_n$, is generated. Where $N_n$ is the number of grid
nodes. This mesh is employed as a support to compute gradients and divergences.

Introducing \eqref{eq:particle_acceleration_forces},
\eqref{eq:particle_internal_forces}, \eqref{eq:particle_body_forces}, 
and \eqref{eq:particle_load_forces} in \eqref{eq:BalanceMomentum_wf},
and approximating the real and virtual velocity  field of the particle
$p$ as
\begin{equation}
  \label{eq:Approx-velocity-mesh}
  \vec{v}_{p} = \sum^{Nn}_I N_I(\vec{x}_p)\cdot \vect{v}_I
  \quad \text{and} \quad \vec{v}^{\psi}_{p} = \sum^{Nn}_I N_I(\vec{x}_p)\cdot \vect{1},
\end{equation}
we reach to the particle balance of forces of the continuum,
\begin{equation}
  \label{eq:particle_balance_forces3}
  \dot{\vec{p}}_{I}= \tens{m}_{IJ}\dot{\vec{v}}_{J} = \vec{f}_{I}^{int} + \vec{f}_{I}^{ext},
\end{equation}
where $\dot{\vec{p}}_{I}$ is the rate of momentum at grid node $I$, the nodal mass matrix $\tens{m}_{IJ}$ is,
\begin{equation}
  \label{eq:particle_nod_mass}
  \tens{m}_{IJ} =
  \sum^{Np}_{p} N_{Ip} m_p N_{Jp}.
\end{equation}
To improve the computational efficiency and stability, the nodal mass matrix
\eqref{eq:particle_nod_mass} can be substituted by the lumped mass
matrix $\tens{m}_{IJ}^{lumped}$.
Later, internal and external forces are computed as follows,
\begin{equation}
  \label{eq:nodal_internal_forces}
  \vec{f}_{I}^{int} = - \sum^{N_p}_{p} \tens{\sigma}_{p} \colon \Grad{N_{Ip}} \frac{m_p}{\rho_p}
\end{equation}
\begin{equation}
  \label{eq:nodal_external_forces}
  \vec{f}_{I}^{ext} = \sum^{N_p}_{p} \vec{b}_{p} N_{Ip} m_p  + \sum^{N_p}_{p}
  N_{Ip} \vec{t}^s_{p} m_p h^{-1} 
\end{equation}
where $\tens{\sigma}_{p} = \tens{\sigma}_{p}(\tens{\varepsilon}_{p})$
is the particle $p$ stress field, which can be integrated employing
the suitable constitutive model. The strain tensor is updated employing the rate of stress tensor $\dot{ \tens{\varepsilon}}_{p}$ used to update the
strain tensor is as follows \eqref{eq:IncrStrainPoint}.
\begin{equation}
  \label{eq:IncrStrainPoint}
  \dot{\tens{\varepsilon}_{p}} = \frac{\Delta
    \tens{\varepsilon}_{p}}{\Delta t} =
  \frac{1}{2} \sum^{Nn}_I \left[\Grad{N_{Ip}}\ \otimes \vec{v}_{I} + \vec{v}_{I} \otimes
    \Grad{N_{Ip}}\ \right].
\end{equation}
Next, imposing $\frac{D \rho}{D t} = 0$, we ensures the mass
conservation and update the density field.
\begin{equation}
  \label{eq:MassConservation}
\dot{\rho} = - \rho\ \mathit{tra} \left( \dot{\tens{\varepsilon}} \right)
\end{equation}

Finally, to solve the equation \eqref{eq:particle_balance_forces3}, a second order
temporal integration scheme is required. Therefore, time is
discretized in to a finite set of time steps $k = 1\ldots ,Nt$, where $k$ is the current time step and $N_t$
is the total number of time steps. Once the nodal equilibrium equation it is solved, the values in the nodes are
interpolated back in to the particles and each particle it is advected
to the new position,
\begin{equation}
  \label{eq:Updated_Lagrangian}
  \dot{\vec{v}}_p = \sum^{N_n}_{I}\vec{a}_{I}\ N_{Ip},\quad and\quad \dot{\vec{x}}_{p} = \sum^{N_n}_{I}\vec{v}_{I}\ N_{Ip}  
\end{equation}
In the MPM literature, this equations
(\ref{eq:particle_balance_forces3}) and (\ref{eq:Updated_Lagrangian}),
are solved with a forward Euler algorithm. In the next section, an
alternative time integration scheme based on the Newmark a-form is proposed.

@@

This scheme has prof to damps out the higher frequency noises
\cite{Tran2019e}. But it can present the same numerical instabilities
as in \eqref{eq:Sulsky-1994-UL-x},\eqref{eq:Sulsky-1994-UL-v} when nodal masses become infinitesimal.
Here we propose a explicit predictor-corrector scheme for MPM based on the Newmark a-form with
$\gamma = 0.5$ and $\beta = 0$ which is the central difference
explicit. This method is devoted to solve a system of equations of type
\begin{equation*}
  \Matrix{M}_{IJ}\ddot{\Vector{d}}_{J} + \Matrix{C}_{IJ}\dot{\Vector{d}}_{J} +
  \Matrix{K}_{IJ}\Vector{d}_{J} = \Vector{F}_{I}
\end{equation*}
Nevertheless it can be applied for an isolated mass, therefore it is
possible to apply this methods successfully in the MPM framework as
was proved by \cite{Tran2019e}. Taking the predictor
definition from the classic literature \cite{Hughes2000} and
calculating particles velocity and acceleration as described in 
\cite{Zhang_book_2016}, we get the following expression for the
\textbf{predictor} stage for both velocity and displacement.
\begin{equation}
  \label{eq:5}
  \vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
  \underbrace{\sum^{N_n}_{I}\frac{\vec{p}_{I}^{k}\
      N_{Ip}^k}{\tens{m}_I^k}}_{\vec{v}_p^{k}} + \frac{1}{2}\Delta
  t^2\ \underbrace{\sum^{N_n}_{I}\frac{\vec{f}_{I}^{k}\
      N_{Ip}^k}{\tens{m}_I^k}}_{\vec{a}_p^{k}}
\end{equation}
\begin{equation}
  \label{eq:Predictor-velocity}
  \vec{v}_p^{k+1} = \vec{v}_p^n + (1 - \gamma)\ \Delta t\
  \underbrace{\sum^{N_n}_{I}\frac{\vec{f}_{I}^{k}\
      N_{Ip}^k}{\tens{m}_I^k}}_{\vec{a}_p^{k}}
\end{equation}

Consequently, the \textbf{corrector} stage for the particles velocity is in the
following way
\begin{equation}
  \vec{v}_{p}^{k+1} = \vec{v}_{p}^{k+1} +
  \gamma\ \Delta t\ \underbrace{\sum^{N_n}_{I}\frac{\vec{f}_{I}^{k+1}\ N_{Ip}^{k+1}}{\tens{m}_I^{k+1}}}_{\vec{a}_p^{k+1}}  
\end{equation}

@@

The previous lines describe the main algorithm without defining an
interpolation technique. A direct consequence of this, is that it is
possible to adopt a width range of them. The MPM has been
successfully applied to a wide range of problems due to its ability to
deal with large strain problems without mesh distorsion issues inherent to
mesh based methods like FE. However, in the simulations made with the
original MPM, there are numerical noises when particles crossing
the cell boundaries. These numerical inaccuracies give rise to the
development of other interpolation techniques. Some of this
alternatives techniques are the generalized interpolation material
point method (GIMP) Bardenhagen \& Kober (2004)\cite{Bardenhagen2004},
the dual domain material point method (DDMP) Zhang {\it et al.}
(2011)\cite{Zhang2011a}, the B-Spline MPM Tran {\it et al.}
(2019)\cite{Tran2019a}, the Conservative Taylor Least
Squares reconstruction Wobbes {\it et al.}
(2018)\cite{E_Wobbes_2018} and more recently the local
maximum-entropy (or local \textit{max-ent}) shape function first
introduced by Arroyo \& Ortiz (2006)\cite{Arroyo2006} has been tested
under the MPM framework by Wobbes {\it et al.}
(2020)\cite{Wobbes2020} where they prof that simulations performed with the maxent basis functions
show considerably more accurate stress approximations for
MPM. Although, in \cite{Wobbes2020} authors does not deep in $\beta$
parameter benefits

@@

In this section, the benchmark proposed by Dyka \& Ingel
(1995)\cite{Dyka1995} is considered to illustrate the
capacity of MPM \cite{Sulsky1994} and OTM
models to avoid tensile instabilities.

@@

The results are provided for the explicit predictor-corrector version of the
MPM scheme (Algorithm \ref{alg:PCE-algorithm}) and OTM algorithms. For MPM, piecewise-
linear, uGIMP and \textit{max-ent} basis functions are employed. The OTM
algorithm is used only with \textit{max-ent} shape functions.

\begin{figure}\sidecaption
  \centering
  \resizebox{0.8\hsize}{!}{
    \input{Figures/Dyka_bar.tex}}
  \caption{Geometrical description of the Dyka \cite{Dyka1995} bar. }
  \label{fig:Dyka_Bar}
\end{figure}

@@

>>>>

\section{Introduction}
\label{intro}
Since the proposal of the Material Point Method (\textbf{MPM}) by Sulsky {\it
  et al.} (1994)\cite{Sulsky1994}, as a generalization of
a particle-in-cell method \cite{Harlow1956} and a fluid implicit
particle (FLIP) method \cite{Brackbill1986}, MPM has been used to
simulate dynamic problems. However, in the simulations made with the
original MPM, there are numerical noises \cite{Tran2019e} when it is faced to solve
classic dynamic benchmarks like the one proposed by Dyka \& Ingel
(1995)\cite{Dyka1995}. These numerical inaccuracies give rise to the
development of other interpolation techniques to overcome this limitation. Some of this
alternatives techniques are the generalized interpolation material
point method (GIMP) Bardenhagen \& Kober (2004)\cite{Bardenhagen2004},
the dual domain material point method (DDMP) Zhang {\it et al.}
(2011)\cite{Zhang2011a}, the B-Spline MPM Tran {\it et al.}
(2019)\cite{Tran2019a}, the Conservative Taylor Least
Squares reconstruction Wobbes {\it et al.}
(2018)\cite{E_Wobbes_2018} and more recently the local
maximum-entropy (or local \textit{max-ent}) shape function first
introduced by Arroyo \& Ortiz (2006)\cite{Arroyo2006} has been tested
under the MPM framework by Wobbes {\it et al.}
(2020)\cite{Wobbes2020}. Alternatively to change the spatial
interpolation technique other researchers have overcome the well known
wiggle that tends to appear in this dynamic simulations by filtering
high-frequencies with an alternative time integration scheme, the
implicit GIMP (iGIMP) Charlton {\it et al.}
(2017)\nocite{Charlton2017} or the generalised-$\alpha$ scheme for the
MPM proposed by Tran \& Solowski (2019)\nocite{Tran2019e}.

The aim of this paper is to mitigate this spurious oscillations by the
employ of a suitable combination of the maximum-entropy (or local
\textit{max-ent}) shape functions, and the proposal of a explicit
predictor-corrector scheme. Local \textit{max-ent} shape functions was
first introduced by \cite{Arroyo2006}, it belongs to the class of convex 
approximation schemes and provides a seamless transition between
finite elements (\textbf{FE}) and mesh-free interpolations. The
approximation scheme is based on a compromise between minimizing the
width of the shape function support and maximizing the information
entropy of the approximation. The local \textit{max-ent} approximation
may be regarded as a regularization, or \textit{thermalization}, of
Delaunay triangulation which effectively resolves the degenerate cases
resulting from the lack of uniqueness of the triangulation. Local
\textit{max-ent} basis functions possess many desirable properties for
meshfree algorithms. First of all, they are entirely defined by the
nodal set and the domain of analysis. They are also non-negative,
satisfy the partition of unity property, and provide an exact
approximation for affine functions \cite{Arroyo2006}. This
approximation scheme has been proof to have a good performance under
the dynamic regime by other researchers like Navas {\it et al.}
(2018)\cite{Navas2018a} and Li {\it et al.} (2012)\cite{Li2012} for
Optimal Transportation Meshfree (\textbf{OTM}) method. On the other
hand, the time integration scheme here proposed is an accommodation of
the Newmark a-form with $\gamma = 0.5$ and $\beta = 0$ which is the central difference
explicit (CD). We have choose this method among other suitable
alternatives as Wilson {\it et al.} (1972)\nocite{Wilson1972} or Chung
\& Hulbert (1993)\nocite{Geranlized_alpha_1993} because it simplicity
and it good performance dealing with solids dynamics problems like those
presented in \cite{Navas2018a}.

@@

\section{Brief note concerning the notation}
\label{sec:Notation}
In what follows, we will adopt the following convention. All the
physical variables involved in this problem are collected in Table
\ref{tab:notation-table}.

\begin{table}
  \centering
  \caption{Physical variables involved in the problem}
  \label{tab:notation-table}
  \begin{tabular}{c c c}
    $\rho$ & Density field & Scalar \\
    $\vect{a}$ & Acceleration field & First order tensor \\
    $\vect{v}$ & Velocity field & First order tensor \\
    $\vect{u}$ & Displacement field & Second order tensor \\
    $\tens{\sigma}$ & Cauchy stress tensor & Second order tensor \\
    $\tens{\varepsilon}$ & Cauchy strain tensor & Second order tensor \\
    $\tens{D}$ & Constitutive tensor & Fourth order tensor\\
  \end{tabular}
\end{table}

Three kind of subscript or superscript are used within paper. The
subscript $\qed_p$ is used to define a particle variable. While the
subscript $\qed_I$ is reserved in this notation for denoting nodal
variables. And finally, the superscript $\qed^{\psi}$ involves a virtual
magnitude. For the operators, the convention is : $\dot{\qed}$ and
$\ddot{\qed}$ for the first and second time derivative, $\qued \otimes
\qued$ means the dyadic operator, $(\cdot)$ and $(\colon)$ means the first
and second contraction of a tensor, $\Div{\qed}$
denotes the divergence operator, and finally $\Grad{\qed}$ and
$\GradS{\qed}$ denotes the gradient and its symmetric part.

\section{Derivation of the MPM}
\label{sec:derivation-mpm}
In this section, we provide an overview of the standard explicit MPM
algorithm as presented by \cite{Sulsky1994}. The method has three main
steps: a variational recovery process to project particle data to the
nodes of the grid, a Eulerian step where balance of momentum is
compute by solving internal and external forces in the nodes, and
finally the advection of the Lagrangian particles. We will first give
a introduction to the governing equations and then follow by an
algorithmic description of the process.

\subsection{Governing equations }
\label{sec:governing-equations}
In the MPM approach a continuum is considered. Let define a
$\varOmega$ domain occupied by an elastic body, and $\partial
\varOmega$ the boundary of the domain, where $\partial
\varOmega = \Gamma_d \bigcup \Gamma_n$ and $\Gamma_d \cap  \Gamma_n =
\emptyset$. Note that $\varGamma_d$ is reserved for Dirichlet boundary
conditions, and  $\varGamma_n$ for the Neumann. The behaviour of the
continuum can be described by a set of governing equations which are
(i) the balance of momentum equation
\begin{equation}
  \label{eq:Balance-momentum}
\rho \vect{a} = \Div{\tens{\sigma}} + \rho \vect{b},
\end{equation}
next (ii) the compatibility equation with ensure that the velocity
field is compatible with the Cauchy strain field,
\begin{equation}
  \label{eq:Compatibility-equation}
  \dot{\tens{\varepsilon}} = \GradS{\vect{v}} = \frac{1}{2}\left(
    \Grad{\vect{v}} +  \GradT{\vect{v}} \right),
\end{equation}
later (iii) the constitutive equation, which relates the rate of Cauchy strain
tensor with the rate of Cauchy stress tensor,
\begin{equation}
  \label{eq:Constitutive-equation}
\dot{\tens{\sigma}} = \tens{D} \colon \dot{\tens{\varepsilon}},
\end{equation}
finally we introduce (iv) the mass conservation, which if obtained by
setting to zero the total derivative of the density field,
\begin{equation}
  \label{eq:Rho-material-derivative}
  \frac{D \rho}{D t} = \dot{\rho} + \rho \Div\vect{v}} = 0.
\end{equation}

@@

\subsection{Galerkin procedure}
\label{sec:Galerkin-procedure}
In order to arrive to a finite set of equations, in contrast with the
FEM, in the MPM a double discretization procedure is performed as we will
describe here below. First, the continuum domain
$\Omega$ is discretized with a finite sum of material points (in the
following particles), each one
represent a part of the discretized domain $\varOmega_p \subset
\varOmega$ with $p = 1,2\ldots ,N_p$ where $N_p$ is the number of
particles. The material point $\vec{x}_p$ is defined at the centroid
of each $\Omega_p$, figure \ref{fig:MPM-discretization}.
\begin{figure}\sidecaption
  \centering
  \resizebox{0.9\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/Discretization_MPM}
  }
  \caption{MPM discretization.}
  \label{fig:MPM-discretization}
\end{figure}

Each material point is assigned with initial values of position,
velocity, mass, volume and stress denoted by $\vec{x}_p$,
$\vec{v}_p$, $m_p$,  $V_p$ and $\tens{\sigma}_p$, but also the
virtual field $\vect{v}^{\psi}_{p}$. So employing the definition of the
material integral, where we recover the Riemann integral definition as
an addition  of a finite set of points. Therefore individual terms in
\eqref{eq:BalanceMomentum_wf} are solved as follows. 
\begin{itemize}
\item Acceleration forces :
\begin{equation}
    \label{eq:particle_acceleration_forces}
    \Integral{3}{\rho\ \vec{a}\ \vect{v}^{\psi}} =& \sum^{N_p}_{p}
   \dot{\vec{v}}_{p}\ \vect{v}^{\psi}_{p}\ m_p.
  \end{equation}\\
\item Internal forces :
  \begin{equation}
    \label{eq:particle_internal_forces}
    \Integral{3}{\tens{\sigma}\ \cdot \Grad{\vec{\psi}}} = \sum^{N_p}_{p}
   \tens{\sigma}_{p}\ \cdot \Grad{\tens{\psi}_{p}}\ V_p.
  \end{equation}\\
\item Body forces :
\begin{equation}
  \label{eq:particle_body_forces}
  \Integral{3}{\rho\ \vec{b}\ \vect{v}^{\psi} } = \sum^{N_p}_{p}
  \vec{b}_{p}\ \vect{v}^{\psi}_p\ m_p.
\end{equation}\\
\item Loads :
\begin{equation}
  \begin{aligned}
    \label{eq:particle_load_forces}
    \Integral{2}{\vec{t}\ \vect{v}^{\psi}} = \Integral{2}{\rho\
      \vec{t}^s\ \vect{v}^{\psi}} = \sum^{N_p}_{p} \vec{t}^s_{p}\ \vect{v}^{\psi}_{p}\ h^{-1}\ m_p ,
  \end{aligned} 
\end{equation}
\end{itemize}
where $h$ is the thickness of the continuum in a 2D case. Here is
where the second discretization procedure appears. A background mesh
composed by a finite set of grid points with coordinates $\vec{x}_I\
, I = 1,2\ldots ,N_n$, is generated. Where $N_n$ is the number of grid
nodes. This mesh is employed as a support to compute gradients and divergences.

Introducing \eqref{eq:particle_acceleration_forces},
\eqref{eq:particle_internal_forces}, \eqref{eq:particle_body_forces}, 
and \eqref{eq:particle_load_forces} in \eqref{eq:BalanceMomentum_wf},
and approximating the real and virtual velocity  field of the particle
$p$ as
\begin{equation}
  \label{eq:Approx-velocity-mesh}
  \vec{v}_{p} = \sum^{Nn}_I N_I(\vec{x}_p) \vect{v}_I
  \quad \text{and} \quad \vec{v}^{\psi}_{p} = \sum^{Nn}_I N_I(\vec{x}_p) \vect{1},
\end{equation}
we reach to the particle balance of forces of the continuum,
\begin{equation}
  \label{eq:particle_balance_forces3}
  \dot{\vec{p}}_{I}= \tens{m}_{IJ}\dot{\vec{v}}_{J} = \vec{f}_{I}^{int} + \vec{f}_{I}^{ext},
\end{equation}
where $\dot{\vec{p}}_{I}$ is the rate of momentum at grid node $I$, the nodal mass matrix $\tens{m}_{IJ}$ is,
\begin{equation}
  \label{eq:particle_nod_mass}
  \tens{m}_{IJ} =
  \sum^{Np}_{p} N_{Ip} m_p N_{Jp}.
\end{equation}
To improve the computational efficiency and stability, the nodal mass matrix
\eqref{eq:particle_nod_mass} can be substituted by the lumped mass
matrix $\tens{m}_{IJ}^{lumped}$.
Later, internal and external forces are computed as follows,
\begin{equation}
  \label{eq:nodal_internal_forces}
  \vec{f}_{I}^{int} = - \sum^{N_p}_{p} \tens{\sigma}_{p} \cdot \Grad{N_{Ip}} \frac{m_p}{\rho_p}
\end{equation}
\begin{equation}
  \label{eq:nodal_external_forces}
  \vec{f}_{I}^{ext} = \sum^{N_p}_{p} N_{Ip} \vec{b}_{p} m_p  + \sum^{N_p}_{p}
  N_{Ip} \vec{t}^s_{p} m_p h^{-1} 
\end{equation}
where $\tens{\sigma}_{p} = \tens{\sigma}_{p}(\tens{\varepsilon}_{p})$
is the particle $p$ stress field, which can be integrated employing
the suitable constitutive model. The strain tensor is updated employing the rate of stress tensor $\dot{ \tens{\varepsilon}}_{p}$ used to update the
strain tensor is as follows \eqref{eq:IncrStrainPoint}.
\begin{equation}
  \label{eq:IncrStrainPoint}
  \dot{\tens{\varepsilon}_{p}} = \frac{\Delta
    \tens{\varepsilon}_{p}}{\Delta t} =
  \frac{1}{2} \sum^{Nn}_I \left[\Grad{N_{Ip}}\ \otimes \vec{v}_{I} + \vec{v}_{I} \otimes
    \Grad{N_{Ip}}\ \right].
\end{equation}
Next, imposing $\frac{D \rho}{D t} = 0$, we ensures the mass
conservation and update the density field.
\begin{equation}
  \label{eq:MassConservation}
\dot{\rho} = - \rho\ \mathit{tra} \left( \dot{\tens{\varepsilon}} \right)
\end{equation}

Finally, to solve the equation \eqref{eq:particle_balance_forces3}, a second order
temporal integration scheme is required. Therefore, time is
discretized in to a finite set of time steps $k = 1\ldots ,Nt$, where $k$ is the current time step and $N_t$
is the total number of time steps. Once the nodal equilibrium equation it is solved, the values in the nodes are
interpolated back in to the particles and each particle it is advected
to the new position,
\begin{equation}
  \label{eq:Updated_Lagrangian}
  \dot{\vec{v}}_p = \sum^{N_n}_{I}N_{Ip}\ \vec{a}_{I},\quad and\quad
  \dot{\vec{x}}_{p} = \sum^{N_n}_{I}N_{Ip}\ \vec{v}_{I}  
\end{equation}
In the MPM literature, this equations
(\ref{eq:particle_balance_forces3}) and (\ref{eq:Updated_Lagrangian}),
are solved with a forward Euler algorithm. In the next section, an
alternative time integration scheme based on the Newmark a-form is proposed.

@@

This scheme has prof to damps out the higher frequency noises
\cite{Tran2019e}. But it can present the same numerical instabilities
as in \eqref{eq:Sulsky-1994-UL-x},\eqref{eq:Sulsky-1994-UL-v} when nodal masses become infinitesimal.
Here we propose a explicit predictor-corrector scheme for MPM based on the Newmark a-form with
$\gamma = 0.5$ and $\beta = 0$ which is the central difference
explicit. This method is devoted to solve a system of equations of type
\begin{equation*}
  \Matrix{M}_{IJ}\ddot{\Vector{d}}_{J} + \Matrix{C}_{IJ}\dot{\Vector{d}}_{J} +
  \Matrix{K}_{IJ}\Vector{d}_{J} = \Vector{F}_{I}
\end{equation*}
Nevertheless it can be applied for an isolated mass, therefore it is
possible to apply this methods successfully in the MPM framework as
was proved by \cite{Tran2019e}. Taking the predictor
definition from the classic literature \cite{Hughes2000} and
calculating particles velocity and acceleration as described in 
\cite{Zhang_book_2016}, we get the following expression for the
\textbf{predictor} stage for both velocity and displacement.
\begin{equation}
  \label{eq:5}
  \vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
  \underbrace{\sum^{N_n}_{I}\frac{N_{Ip}^k\ \vec{p}_{I}^{k}}{\tens{m}_I^k}}_{\vec{v}_p^{k}} + \frac{1}{2}\Delta
  t^2\ \underbrace{\sum^{N_n}_{I}\frac{N_{Ip}^k\ \vec{f}_{I}^{k}}{\tens{m}_I^k}}_{\vec{a}_p^{k}}
\end{equation}
\begin{equation}
  \label{eq:Predictor-velocity}
  \vec{v}_p^{k+1} = \vec{v}_p^n + (1 - \gamma)\ \Delta t\
  \underbrace{\sum^{N_n}_{I}\frac{N_{Ip}^k\ \vec{f}_{I}^{k}}{\tens{m}_I^k}}_{\vec{a}_p^{k}}
\end{equation}

Consequently, the \textbf{corrector} stage for the particles velocity is in the
following way
\begin{equation}
  \vec{v}_{p}^{k+1} = \vec{v}_{p}^{k+1} +
  \gamma\ \Delta t\ \underbrace{\sum^{N_n}_{I}\frac{N_{Ip}^{k+1}\ \vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}}_{\vec{a}_p^{k+1}}  
\end{equation}

@@

The previous lines describe the main algorithm without defining an
interpolation technique. A direct consequence of this, is that it is
possible to adopt a width range of them. The MPM has been
successfully applied to a wide range of problems due to its ability to
deal with large strain problems without mesh distorsion issues inherent to
mesh based methods like FE. However, in the simulations made with the
original MPM, there are numerical noises when particles crossing
the cell boundaries. Recently the local maximum-entropy (or local
\textit{max-ent}) shape function first introduced by Arroyo \& Ortiz
(2006)\cite{Arroyo2006} has been tested under the MPM framework by
Wobbes {\it et al.} (2020)\cite{Wobbes2020} where they prof that
simulations performed with the maxent basis functions show
considerably more accurate stress approximations for MPM. Although, in
\cite{Wobbes2020} authors does not deep in $\beta$ parameter
benefits. In this publication a parametric analysis of $\Beta$ will be
performed in order to reach a compromise value that minimize the
presence of wiggle without introducing an additional spurious behaviour.

@@

In this section, the benchmark proposed by Dyka \& Ingel
(1995)\cite{Dyka1995} is considered to illustrate the
capacity of MPM \cite{Sulsky1994} and OTM
models to avoid tensile instabilities. Here we employ our in-house
codes for both MPM and OTM simulations.

@@

The results are provided for the explicit predictor-corrector version
of the MPM scheme (Algorithm \ref{alg:PCE-algorithm}) and OTM
algorithms. For MPM, piecewise-linear, uGIMP and \textit{max-ent}
basis functions are employed. The OTM algorithm is used only with
\textit{max-ent} shape functions.

\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \input{Figures/Dyka_bar.tex}}
  \caption{Geometrical description of the Dyka \cite{Dyka1995} bar.}
  \label{fig:Dyka_Bar}
\end{figure}

@@

The total simulation time is 0.0002 seconds, a direct consequence of
that is we let the elastic wave generated travel thorough the bar
(from A to C and back to A) at least four times. For the spatial
discretization we will set five different values of the it : 0.1,
0.05, 0.01, 0.005 and 0.001 millimeters. The initial mesh will be
composed by a single particle in the center of the element. Regarding
this, here we give one consideration to properly modellize this
benchmark under the MPM framework. Notice that the free border of the
bar has a horizontal displacement of 0.033 millimeters, therefore a
computational domain with an extra gap of 0.033 millimeters is required in
order to accommodate the unconstrained displacement of the particles in the
left border of the bar. Here below the time step is controlled by the courant number, in
this case, we adopt a middle value of it (CFL=0.5) as a compromise
between stability and numerical accuracy.
